--- 
title: "musings of the Professor"
author: "Tehran J. Davis"
date: "`r Sys.Date()`"
site: bookdown::gitbook
documentclass: book
bibliography: [book.bib, packages.bib]
biblio-style: apalike
link-citations: yes
description: "Weekly examples and musings related to PSYC 7014."
---

# Weekly musings {-}

In these (roughly) weekly musings I will supplimenting the course texts and in-class examples with some written walk-thoughs. In previous years students have informed me that they got a lot out of these little jaunts. My hope is that they will be of some use to you.

If you want to follow along you can copy and paste the relevant code chunks into your own Rmd file and execute.

Note that in what follows code chunks and outputs are in blockquotes with the colored tint. Outputs that are generated by the code have hash marks in front of them.

```{r eval=FALSE, echo=FALSE}
install.packages("bookdown")
# or the development version
# devtools::install_github("rstudio/bookdown")
```

```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown'
), 'packages.bib')
```

```{r setup, include=FALSE, echo=FALSE}
r <- getOption("repos")
r["CRAN"] <- "http://cran.cnr.berkeley.edu/"
options(repos = r)
require(knitr)
# Set so that long lines in R will be wrapped:
opts_chunk$set(tidy.opts=list(width.cutoff=80),tidy=TRUE)
```